---
title: 'Coronavirus: An Effective Campaign Strategy'
author: Samuel Lyubic
date: '2020-09-04'
slug: 5523-blog-one
categories:
  - Statistics
  - Politics
tags:
  - Serbia
  - Coronavirus
subtitle: 'An analysis into how coronavirus may have impacted Serbias election'
summary: 'An analysis into how coronavirus may have impacted Serbias election'
authors: []
lastmod: '2020-09-04T02:20:49+10:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "centre")
library(coronavirus)
library(tidycovid19)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(DT)
library(widgetframe)
library(kableExtra)
library(httpuv)
#update_dataset()
```

```{r read-in-data}
covid_df <- download_merged_data(cached = TRUE, silent = TRUE) %>%
  filter(iso3c == "SRB")
```

# Introduction

Coronavirus was labelled a pandemic on the 11th of March and since then has wreaked havoc across the world, leaving individuals, economies and governments scrambling to find a way through this global event. There have been a range of measures and policies issued by governments across the world to try combat and overcome the devastation that coronavirus has inflicted. Many of these measures involve social distancing, which can involve restricting the number of people that can be within a certain radius of each other as well as the number of people that can collectively be in a public space together, as such many proceedings requiring people to gather are now considered illegal and a risk to public health and safety resulting in their postponement or cancellation. An event such as an election would usually not have been considered it was safe or not, is now unable to go forward as, in order for an election to take place the citizens of a country are required to vote and in order to vote, the citizens need to congregate, usually there would be nothing unique about this statement however, given the global event that world has been stricken with, it would be seem an election would be a hard thing to pull off, that is, unless you’re Serbia. 

Serbia just recently held a parliamentary election on June 21st 2020 in the heart of the global pandemic that is the Coronavirus. Many citizens, pundits and opposing parties have raised concerns over the legitimacy of the election as many people believe it was wrought with the feeling of an authoritarian regime rather than a democratic election, given the option to vote were very limited due to the parliamentary boycott that took place whereby the Serbia Progressive Party (SNS) almost ran unopposed as their opposition boycotted the election. Given the boycot that took place, many reports indicate that the SNS were looking to have the election as soon as possible leading to falsified reports of the coronavirus impact and putting the nation's health and safety at risk for the personal gain of the party[https://balkaninsight.com/2020/07/17/vucic-is-paying-a-price-for-serbias-bogus-elections/], which indicates coronavirus has had a significant impact on the election and this potential link will be further explored to assess how coronavirus may have assisted the SNS party into re-election.


# The Data

The data used in the analysis has been sourced from the _tidycovid19_ package which has been developed by Joachim Gassen [https://github.com/joachim-gassen/tidycovid19] with data being pulled from:
  - The John Hopkins University Center for Systems Science and Engineering, 
  - European Centre for Disease Prevention and Control
  - Our World in Data
  - World Bank
  - ACAPS, 
  - Apple Maps 
  - Oxford OxCGRT team
  
The key variables being:
  - Confirmed: Which is the accumualted confirmed coronavirus cases
  - Deaths: The accumulated conornavirus deaths
  - Soc_dist: The number of social distancing measures reported up to date by ACAPS, net of lifted restrictions

For the analysis to take place, the following variables were created:
  - Growth_factor: A measure of how fast the numbers of a new cases are going up or down. Which is calculated by taking the daily reported cases and dividing them by the average of the reported cases for the previous five days. by  A growth factor above 1 indicates the numbers are going up and a growth factor below 1 indicates the virus is getting under control. [https://www.abc.net.au/news/2020-04-10/coronavirus-data-australia-growth-factor-covid-19/12132478?nw=0].
  - Change_cases%: Measures the percent change in confirmed daily cases from the previous day.  
  - Change_deaths%: Measures the percent change in confirmed deaths from the previous day.
 
```{r select-variables}
covid_srb <- covid_df %>%
  select(date, confirmed, deaths, soc_dist, mov_rest) %>%
  mutate(month = month(date, labe = TRUE, abbr = TRUE),
         lag_confirmed = lag(confirmed),
         lag_deaths = lag(deaths),
         daily_cases = confirmed - lag_confirmed,
         daily_deaths = deaths - lag_deaths,
         soc_dist = as.factor(soc_dist),
         mov_rest = as.factor(mov_rest)) %>%
  filter(month >= "Feb") %>% 
  select(-lag_confirmed, -lag_deaths)
```

# Serbia reacts to coronavirus

Prior to being labelled a pandemic, the novel coronavirus had already run through countries like China and Italy causing mass devesation giving the virus global attention. On February 26th one of the presidents, Aleksandar Vucic, advisors making light of the virus referring to it as "the funniest virus" [https://balkaninsight.com/2020/04/03/serbia-pins-coronavirus-blame-on-returning-serbs-concealing-infection/] subsequently, the virus was labelled a pandemic on teh 11th of March and the Serbian president declared a state of emergency on the 15th of March [https://www.srbija.gov.rs/vest/en/151398/state-of-emergency-declared-throughout-serbia.php]. Figure \@ref(fig:plot-daily-cases-deaths) visualises both the counts for the daily confirmed cases and daily deaths for Serbia, with cases and deaths along the y axis and the date on the x axis. The verticle red lines indicate declerations of state of emergencies, the orange line indicates the state of emergency being called off and the blue line indicates the serbian election. It is evident that Serbia saw a steep accumulation of cases and deaths from the middle of May through to the middle of April and then a somewhat declining trend of cases and deaths from may through to the end of July where there was again a signifcant increase in cases and deaths. 


```{r plot-daily-cases-deaths, fig.cap = "Visualises the daily counts for confirmed cases and deaths"}
cases_plot <- covid_srb %>% 
  filter(date >= "2020-03-01") %>%
  ggplot(aes(x = date,
             y = daily_cases)) +
  geom_col(stat = "identity", 
           alpha = .6,
           fill = "#4545ff") +
  ggtitle("Daily Confirmed Cases") +
  xlab("Date") +
  ylab("Daily Cases") +
  theme_linedraw() +
  scale_x_date(breaks = 'month',
               date_labels = "%b %d") +
  geom_vline(xintercept = as.numeric(as.Date("2020-03-15")), color = "red") +
  geom_vline(xintercept = as.numeric(as.Date("2020-05-06")), color = "orange") +
  geom_vline(xintercept = as.numeric(as.Date("2020-06-21")), color = "#004080") +
  geom_vline(xintercept = as.numeric(as.Date("2020-07-03")), color = "red") 


deaths_plot <- covid_srb %>% 
  filter(date >= "2020-03-01") %>%
  ggplot(aes(x = date,
           y = daily_deaths)) +
  geom_col(stat = "identity", 
           alpha = .6,
           fill = "#404040") +
  ggtitle("Daily Deaths") +
  xlab("Date") +
  ylab("Daily Deaths") +
  theme_linedraw() +
  scale_x_date(breaks = 'month',
               date_labels = "%b %d") +
  geom_vline(xintercept = as.numeric(as.Date("2020-03-15")), color = "red") +
  geom_vline(xintercept = as.numeric(as.Date("2020-05-06")), color = "orange") +
  geom_vline(xintercept = as.numeric(as.Date("2020-06-21")), color = "#004080") +
  geom_vline(xintercept = as.numeric(as.Date("2020-07-03")), color = "red") 

grid.arrange(cases_plot, deaths_plot, nrow = 2)
```

```{r summary-statistics}
#finding the average of the last five days, every day
covid_growth_rate <- covid_srb %>%
  mutate((five_day = daily_cases + lag(daily_cases) + lag(daily_cases, n = 2) + lag(daily_cases, n = 3) + lag(daily_cases, n = 4))/5,
         growth_rate = (five_day/lag(five_day)),
         growth_rate = round((growth_rate), 2),
         pct_change_cases = round((((daily_cases/lag(daily_cases)) - 1)*100), 2),
         pct_change_deaths = round((((daily_deaths/lag(daily_deaths)) - 1)*100), 2)) %>%
  replace_na(list(growth_rate = 0,
                  pct_change_cases = 0,
                  pct_change_deaths = 0,
                  five_day_ave  = 0)) %>%
  filter(month < "Sep") 

covid_summary <- covid_growth_rate %>%
  filter(month >= "Mar") %>%
  select(date, growth_rate, daily_cases, pct_change_cases, daily_deaths, pct_change_deaths) %>%
  rename(`change_deaths%` = pct_change_deaths,
         `change_cases%` = pct_change_cases,
         growth_factor = growth_rate)

  
summary_table <- covid_summary %>%
  datatable(class = 'cell-border stripe', rownames = FALSE, extensions = 'FixedColumns', options = list(
  pageLength = 5,
  lengthMenu = c(5, 10, 15, 20),
  scrollX = TRUE,
  scrollCollapse = TRUE))

covid_average_summary <- covid_growth_rate %>%
  mutate(pct_change_cases = replace(pct_change_cases, is.infinite(pct_change_cases), NA),
         pct_change_deaths = replace(pct_change_deaths, is.infinite(pct_change_deaths), NA),
         growth_rate = replace(growth_rate, is.infinite(growth_rate), NA)) %>%
  replace_na(list(growth_rate = 0,
                  pct_change_cases = 0,
                  pct_change_deaths = 0,
                  five_day_ave  = 0)) %>%
  filter(month >= "Mar") %>%
  select(date, month, growth_rate, daily_cases, pct_change_cases, daily_deaths, pct_change_deaths) %>%
  rename(`change_deaths%` = pct_change_deaths,
         `change_cases%` = pct_change_cases,
         growth_factor = growth_rate)

table_averages_summary <- covid_average_summary %>%
  group_by(month) %>%
  summarise(`Growth_Factor_Ave` = round((mean(growth_factor )), 3),
            `Daily_Cases%_Ave` = round((mean(`change_cases%`)),3),
            `daily_deaths%_ave` = round((mean(`change_deaths%`)), 3)) %>%
  rename(Month = month) %>%
  select(Month,`Growth_Factor_Ave`, `Daily_Cases%_Ave`)

ave_table <- table_averages_summary %>%
datatable(class = 'cell-border stripe', rownames = FALSE, extensions = 'FixedColumns', options = list(
  pageLength = 5,
  lengthMenu = c(5, 10, 15, 20),
  scrollX = TRUE,
  scrollCollapse = TRUE))
```



```{r proportion-distancing-measure}
dist_monthly_proportion <- covid_srb %>%
  group_by(month, soc_dist) %>%
  summarise(prop = n()) %>%
  mutate(proportion = round(((prop/sum(prop))*100), 2)) %>%
  ungroup() %>%
  select(month, soc_dist, proportion) %>%
  pivot_wider(names_from = "soc_dist",
              values_from = "proportion") %>%
  mutate_all(~replace(., is.na(.), 0)) %>%
  rename(Month = month,
         Zero = `0`,
         One = `1`,
         Two = `2`,
         Three = `3`,
         Four = `4`,
         Five = `5`,
         Six = `6`) %>%
  arrange(Month, Zero, One, Two, Three, Four, Five, Six)

measure_pct <- dist_monthly_proportion %>% 
  datatable(class = 'cell-border stripe', rownames = FALSE, options = list(
  pageLength = 5,
  lengthMenu = c(5, 10, 15, 20)))
```

```{css, echo = FALSE}
dataTables_ScrollHeadInner { padding-right : 10px; }
```


```{r summary-table-widget, fig.cap = "Table displaying covid19 summary statistics for Serbia"}
frameWidget(summary_table)
```

Table \@ref(fig:summary-table-widget) presents the growth factor, daily confirmed cases and deaths and percent change in daily confirmed cases and deaths since March, with Table \@ref(fig:summary-ave-widget) showing the monthly averages for those statistics. The recorded daily growth rate, percentage change in daily cases and deaths and daily cases and death count can be viewed through Table \@ref(fig:summary-table-widget). By looking at Table \@ref(fig:summary-ave-widget) the averages for those statistics can viewed, which shows that in March and April the daily confirmed cases on average as growing and as growth factor was 1.129 and 1.036 showing the growth factor was consistently above 1 for two months straight andt the virus was not undercontrol. This resulting in social distancing measures to increase across March and April, as shown by Table \@ref(fig:measure-widget-plot) march was split between zero and one measure while 86.33% of April was spent with six social distancing restrictions in place  




```{r summary-ave-widget, fig.cap = "Average of the key statistics"}
frameWidget(ave_table)
```






```{r measure-widget-plot, fig.cap = "Table displaying the proportion each measure has been used in each month"}
frameWidget(measure_pct)
```



This will be a conclusion. 


